---
title: "STA437 Midterm Project — Human Activity Recognition (HAR)"
author: "Gaby"
date: "March 2026"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    highlight: kate
    fig_width: 10
    fig_height: 7
    df_print: paged
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  cache   = FALSE,   # never cache — ensures full re-execution on each knit
  fig.path = "figures/",
  dpi      = 150
)
```

> **Dataset:** `har_data.csv` — 10,299 observations, 33 numeric features, 6 activity classes
> **AI Attribution:** Claude (Anthropic) was used to help plan the analysis workflow, suggest diagnostic checks, and improve interpretation of PCA/FA/ICA outputs. All code was reviewed, executed, and validated by the author.

---

# Setup and Package Imports

```{r packages}
# ── Install packages if needed (run once) ────────────────────────────────────
# install.packages(c("tidyverse", "corrplot", "ggplot2", "psych",
#                    "GPArotation", "FactoMineR", "factoextra",
#                    "fastICA", "CCA", "MVN", "gridExtra", "viridis",
#                    "RColorBrewer", "kableExtra", "patchwork", "MASS"))

# ── Load packages ─────────────────────────────────────────────────────────────
library(tidyverse)      # dplyr, ggplot2, readr, tidyr
library(corrplot)       # correlation heatmap
library(psych)          # fa(), KMO(), cortest.bartlett(), principal()
library(GPArotation)    # varimax / promax rotations for psych::fa()
library(FactoMineR)     # PCA()
library(factoextra)     # fviz_eig(), fviz_pca_ind()
library(fastICA)        # fastICA()
library(CCA)            # cc()
library(MVN)            # mvn() — Mardia / HZ tests
library(gridExtra)      # grid.arrange()
library(viridis)        # colour scales
library(RColorBrewer)   # palettes
library(kableExtra)     # kable() formatting
library(patchwork)      # combine ggplot panels
library(MASS)           # ginv() — Moore-Penrose pseudoinverse

# ── Reproducibility ───────────────────────────────────────────────────────────
SEED <- 437
RNGkind("Mersenne-Twister", "Rejection")  # R >= 3.6 default; explicit for portability
set.seed(SEED)

# ── Output directories ────────────────────────────────────────────────────────
dir.create("figures", showWarnings = FALSE)
dir.create("output",  showWarnings = FALSE)

cat("All packages loaded. SEED =", SEED, "\n")
```

---

# Load Data

```{r load-data}
DATA_PATH <- "har_data.csv"
har       <- read_csv(DATA_PATH, show_col_types = FALSE)

cat("Shape:", nrow(har), "×", ncol(har), "\n")
cat("Column names:\n")
print(names(har))

head(har, 3)
```

## Data Audit

```{r data-audit}
cat("=== Data Audit ===\n")
cat("Rows       :", nrow(har), "\n")
cat("Columns    :", ncol(har), "\n")
cat("Missing    :", sum(is.na(har)), "\n")
cat("Duplicates :", sum(duplicated(har)), "\n\n")

cat("Activity distribution:\n")
print(table(har$activity))
```

> **Data audit summary:**
> - n = 10,299 observations, p = 33 numeric features + 1 activity label
> - No missing values, no duplicate rows
> - All numeric variables are already normalised to [−1, 1] per the dataset README
> - `activity` is the categorical label; retained for colour-coding visualisations but **excluded from multivariate analysis**

---

# Data Cleaning and Preprocessing

```{r preprocessing}
# ── Separate label and features ───────────────────────────────────────────────
y     <- factor(har$activity)
X_raw <- har %>% select(-activity)

feature_names <- names(X_raw)

# Abbreviated names for plot readability
short_names <- feature_names %>%
  str_replace_all("Body",    "B")   %>%
  str_replace_all("Gravity", "Grav") %>%
  str_replace_all("mean\\(\\)", "m") %>%
  str_replace_all("Mag",    "Mg")  %>%
  str_replace_all("Acc",    "A")   %>%
  str_replace_all("Gyro",   "G")   %>%
  str_replace_all("Jerk",   "J")

cat("Feature matrix :", nrow(X_raw), "×", ncol(X_raw), "\n")
cat("Label vector   :", length(y), "\n")
cat("Classes        :", levels(y), "\n")
```

```{r standardise}
# ── Standardise (z-score) — PCA/FA based on correlation matrix ────────────────
X_scaled <- scale(X_raw)          # n × p matrix, mean=0, sd=1

cat("After standardisation:\n")
cat("  mean ≈", round(mean(X_scaled), 6), "(expect 0)\n")
cat("  sd   ≈", round(mean(apply(X_scaled, 2, sd)), 6), "(expect 1)\n")

X_df <- as.data.frame(X_scaled)
colnames(X_df) <- feature_names

# Save
write.csv(X_df, "output/X_scaled.csv", row.names = FALSE)
cat("Saved X_scaled.csv\n")
```

---

# Exploratory Data Analysis (EDA)

## Summary Statistics

```{r summary-stats}
desc <- X_df %>%
  pivot_longer(everything(), names_to = "feature", values_to = "value") %>%
  group_by(feature) %>%
  summarise(
    mean   = round(mean(value),   3),
    sd     = round(sd(value),     3),
    min    = round(min(value),    3),
    q25    = round(quantile(value, 0.25), 3),
    median = round(median(value), 3),
    q75    = round(quantile(value, 0.75), 3),
    max    = round(max(value),    3),
    .groups = "drop"
  )

write.csv(desc, "output/summary_stats.csv", row.names = FALSE)

desc %>%
  kable(caption = "Summary Statistics (standardised features)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                font_size = 11, full_width = FALSE)
```

## Distribution Panel

```{r fig1-distributions, fig.width=11, fig.height=8, fig.cap="Distributions of 9 representative features by activity"}
# Pick 9 evenly-spaced features
idx9 <- round(seq(1, 33, length.out = 9))
sel_features <- feature_names[idx9]
sel_short    <- short_names[idx9]
act_palette  <- brewer.pal(6, "Set2")
activity_list <- levels(y)

plot_list <- map2(sel_features, sel_short, function(col, sname) {
  df_plot <- data.frame(value = X_df[[col]], activity = y)
  ggplot(df_plot, aes(x = value, fill = activity, colour = activity)) +
    geom_density(alpha = 0.4, linewidth = 0.4) +
    scale_fill_brewer(palette = "Set2") +
    scale_colour_brewer(palette = "Set2") +
    labs(title = sname, x = "Standardised value", y = "Density") +
    theme_minimal(base_size = 8) +
    theme(legend.position = "none",
          plot.title = element_text(size = 8, face = "bold"))
})

# Combine panels
legend_plot <- ggplot(data.frame(activity = y, v = 1),
                      aes(x = v, fill = activity)) +
  geom_bar() +
  scale_fill_brewer(palette = "Set2") +
  theme_void() +
  theme(legend.position = "bottom",
        legend.title = element_text(size = 9),
        legend.text  = element_text(size = 8))

fig1 <- wrap_plots(plot_list, ncol = 3) +
  plot_annotation(
    title = "Distributions of 9 Representative Features by Activity",
    theme = theme(plot.title = element_text(size = 11, face = "bold"))
  )

# Save
ggsave("figures/fig1_distributions.png", fig1, width = 11, height = 8, dpi = 150)
fig1
```

## Correlation Heatmap

```{r fig2-correlation, fig.width=11, fig.height=10, fig.cap="Feature correlation matrix (lower triangle)"}
corr_mat <- cor(X_df)

# Save heatmap via corrplot (base R graphics)
png("figures/fig2_correlation_heatmap.png", width = 1600, height = 1400, res = 150)
corrplot(corr_mat,
         method   = "color",
         type     = "lower",
         col      = colorRampPalette(c("firebrick3", "white", "steelblue4"))(200),
         tl.cex   = 0.55,
         tl.col   = "black",
         tl.srt   = 90,
         addgrid.col = "gray90",
         diag     = FALSE,
         cl.cex   = 0.7,
         title    = "Feature Correlation Matrix (Lower Triangle)",
         mar      = c(0, 0, 2, 0))
dev.off()

corrplot(corr_mat,
         method   = "color",
         type     = "lower",
         col      = colorRampPalette(c("firebrick3", "white", "steelblue4"))(200),
         tl.cex   = 0.55,
         tl.col   = "black",
         tl.srt   = 90,
         addgrid.col = "gray90",
         diag     = FALSE,
         cl.cex   = 0.7,
         title    = "Feature Correlation Matrix (Lower Triangle)",
         mar      = c(0, 0, 2, 0))

# Count high-correlation pairs
high_corr_pairs <- sum(abs(corr_mat[lower.tri(corr_mat)]) > 0.80)
cat("\nPairs with |r| > 0.80:", high_corr_pairs, "\n")
```

> **EDA takeaway:** The correlation heatmap reveals clear block structure — time-domain accelerometer features cluster together, as do frequency-domain features. Multiple pairs show |r| > 0.80, motivating a latent-dimension approach.

---

# Assumption Testing

## Multivariate Normality — Mardia's Test & QQ-plot

```{r fig3-mvn, fig.width=6, fig.height=5, fig.cap="MVN assessment: Mahalanobis distance QQ-plot"}
# Subsample n = 1000 for speed
set.seed(SEED)
idx_sub <- sample(nrow(X_scaled), 1000)
X_sub   <- X_scaled[idx_sub, ]

# Mahalanobis distances
mu_sub  <- colMeans(X_sub)
cov_sub <- cov(X_sub)
# 协方差矩阵可能近奇异（有完全相关的列），用伪逆替代普通逆
cov_inv <- MASS::ginv(cov_sub)
D2      <- mahalanobis(X_sub, mu_sub, cov_inv, inverted = TRUE)

p_dim   <- ncol(X_scaled)
theo_q  <- qchisq(ppoints(length(D2)), df = p_dim)

png("figures/fig3_mvn_qqplot.png", width = 900, height = 750, res = 150)
qqplot(theo_q, sort(D2),
       xlab = paste0("Chi²(", p_dim, ") quantiles"),
       ylab = "Mahalanobis² quantiles",
       main = "MVN Assessment: Mahalanobis Distance QQ-plot\n(n=1000 subsample)",
       pch = 20, cex = 0.5, col = "steelblue")
abline(0, 1, col = "firebrick", lty = 2, lwd = 2)
legend("topleft", legend = "MVN reference line",
       col = "firebrick", lty = 2, lwd = 2, cex = 0.8)
dev.off()

qqplot(theo_q, sort(D2),
       xlab = paste0("Chi²(", p_dim, ") quantiles"),
       ylab = "Mahalanobis² quantiles",
       main = "MVN Assessment: Mahalanobis Distance QQ-plot\n(n=1000 subsample)",
       pch = 20, cex = 0.5, col = "steelblue")
abline(0, 1, col = "firebrick", lty = 2, lwd = 2)
legend("topleft", legend = "MVN reference line",
       col = "firebrick", lty = 2, lwd = 2, cex = 0.8)
```

```{r mvn-mardia}
# Mardia's test (skewness + kurtosis) via MVN package
# Uses a subsample for speed; adjust size if needed
mvn_result <- mvn(data             = X_sub,
                  test             = "mardia",
                  univTest         = "SW",
                  multivariatePlot = "none")

cat("=== Mardia's Multivariate Normality Test ===\n")
print(mvn_result$multivariateNormality)

cat("\n=== Shapiro-Wilk marginal normality (n=200 subsample) ===\n")
set.seed(SEED)
idx200  <- sample(nrow(X_scaled), 200)
sw_pvals <- apply(X_scaled[idx200, ], 2, function(v) shapiro.test(v)$p.value)
n_normal <- sum(sw_pvals > 0.05)
cat(sprintf("%d / %d features not rejected at α=0.05\n", n_normal, p_dim))
cat("  => Marginal normality:",
    ifelse(n_normal > p_dim * 0.7, "generally supported", "questionable"), "\n")
```

## KMO Test

```{r kmo}
kmo_result <- KMO(X_df)

cat(sprintf("KMO overall: %.4f\n", kmo_result$MSA))
interp <- dplyr::case_when(
  kmo_result$MSA >= 0.90 ~ "marvellous",
  kmo_result$MSA >= 0.80 ~ "meritorious",
  kmo_result$MSA >= 0.70 ~ "middling",
  kmo_result$MSA >= 0.60 ~ "mediocre",
  TRUE                   ~ "miserable"
)
cat("  → KMO interpretation:", interp, "\n\n")

# Per-variable KMO
kmo_per_var <- data.frame(
  feature = short_names,
  KMO     = round(kmo_result$MSAi, 4)
) %>% arrange(KMO)

write.csv(kmo_per_var, "output/kmo_per_variable.csv", row.names = FALSE)
cat("Min KMO per variable:", min(kmo_per_var$KMO), "| Max:", max(kmo_per_var$KMO), "\n")

kmo_per_var %>%
  kable(caption = "Per-variable KMO values") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"),
                font_size = 11, full_width = FALSE)
```

## Bartlett's Test of Sphericity

```{r bartlett}
bart_result <- cortest.bartlett(corr_mat, n = nrow(X_df))

cat(sprintf("Bartlett's test: χ²(%d) = %.2f,  p = %.2e\n",
            bart_result$df, bart_result$chisq, bart_result$p.value))
cat("  → Correlation matrix significantly differs from identity:",
    ifelse(bart_result$p.value < 0.05, "YES", "NO"), "\n")
```

## Extreme Multicollinearity Check

```{r multicollinearity}
corr_upper  <- corr_mat
corr_upper[lower.tri(corr_upper, diag = TRUE)] <- NA
extreme_idx <- which(abs(corr_upper) > 0.95, arr.ind = TRUE)

cat(sprintf("Pairs with |r| > 0.95: %d\n", nrow(extreme_idx)))
if (nrow(extreme_idx) > 0) {
  for (i in seq_len(nrow(extreme_idx))) {
    r_val <- round(corr_upper[extreme_idx[i, 1], extreme_idx[i, 2]], 4)
    cat(sprintf("  %s  ↔  %s  |  r = %.4f\n",
                short_names[extreme_idx[i, 1]],
                short_names[extreme_idx[i, 2]], r_val))
  }
}
```

> **Assumption testing summary:**
> - **MVN:** QQ-plot shows systematic departure from chi²(33) — data are not MVN. This motivates ICA (which exploits non-Gaussianity).
> - **KMO:** ≥ 0.80 strongly supports factor analysis.
> - **Bartlett:** Significant (p ≪ 0.05) → FA is appropriate.
> - **Multicollinearity:** Small number of near-duplicate pairs; FA naturally absorbs collinear variables into shared factors.

---

# Principal Component Analysis (PCA)

```{r pca-full}
# ── Full PCA via prcomp (correlation matrix = scale=TRUE; already scaled here)
pca_full <- prcomp(X_scaled, center = FALSE, scale. = FALSE)

eigvals   <- pca_full$sdev^2
var_ratio <- eigvals / sum(eigvals)
cum_var   <- cumsum(var_ratio)

# PCs explaining 80%, 90%, 95%
for (thresh in c(0.80, 0.90, 0.95)) {
  k <- which(cum_var >= thresh)[1]
  cat(sprintf("%d%% variance explained by %d PCs\n", as.integer(thresh*100), k))
}

# Save eigenvalue table
eig_table <- data.frame(
  PC          = 1:p_dim,
  eigenvalue  = round(eigvals,   4),
  var_explained = round(var_ratio, 4),
  cumulative  = round(cum_var,   4)
)
write.csv(eig_table, "output/pca_eigenvalues.csv", row.names = FALSE)

cat("\nFirst 10 PCs:\n")
eig_table[1:10, ] %>%
  kable(caption = "PCA Eigenvalue Table (first 10 PCs)") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"),
                font_size = 11, full_width = FALSE)
```

## Scree Plot

```{r fig4-scree, fig.width=12, fig.height=5, fig.cap="PCA scree plot and cumulative variance"}
n_show <- 20

df_eig <- data.frame(
  PC      = 1:n_show,
  eigval  = eigvals[1:n_show],
  cumvar  = cum_var[1:n_show] * 100
)

p_scree <- ggplot(df_eig, aes(x = PC, y = eigval)) +
  geom_line(colour = "steelblue", linewidth = 1) +
  geom_point(colour = "steelblue", size = 2.5) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "firebrick",
             linewidth = 0.9) +
  annotate("text", x = n_show - 2, y = 1.15, label = "Kaiser (λ=1)",
           colour = "firebrick", size = 3.5) +
  scale_x_continuous(breaks = 1:n_show) +
  labs(title = "Scree Plot", x = "Principal Component", y = "Eigenvalue") +
  theme_minimal(base_size = 11)

p_cumvar <- ggplot(df_eig, aes(x = PC, y = cumvar)) +
  geom_line(colour = "darkorange", linewidth = 1) +
  geom_point(colour = "darkorange", size = 2.5) +
  geom_hline(yintercept = c(80, 90, 95),
             linetype   = c("dotted","dotted","dotted"),
             colour     = c("gray50", "darkgreen", "purple"),
             linewidth  = 0.9) +
  annotate("text", x = rep(n_show - 1, 3), y = c(81.5, 91.5, 96.5),
           label  = c("80%", "90%", "95%"),
           colour = c("gray50","darkgreen","purple"), size = 3) +
  scale_x_continuous(breaks = 1:n_show) +
  scale_y_continuous(limits = c(0, 105)) +
  labs(title = "Cumulative Variance Explained",
       x = "Number of PCs", y = "Cumulative Variance (%)") +
  theme_minimal(base_size = 11)

fig4 <- p_scree + p_cumvar +
  plot_annotation(title = "PCA: Scree Plot and Cumulative Variance",
                  theme = theme(plot.title = element_text(size = 12, face = "bold")))

ggsave("figures/fig4_pca_scree.png", fig4, width = 12, height = 5, dpi = 150)
fig4
```

## PC Score Plot (PC1 vs PC2)

```{r fig5-scores, fig.width=8, fig.height=6, fig.cap="PCA score plot: PC1 vs PC2 coloured by activity"}
scores <- pca_full$x

df_scores <- data.frame(
  PC1      = scores[, 1],
  PC2      = scores[, 2],
  activity = y
)

fig5 <- ggplot(df_scores, aes(x = PC1, y = PC2, colour = activity)) +
  geom_point(size = 0.8, alpha = 0.35) +
  scale_colour_brewer(palette = "Set2") +
  labs(
    title  = "PCA Score Plot: PC1 vs PC2 (coloured by activity)",
    x      = sprintf("PC1 (%.1f%% var)", var_ratio[1] * 100),
    y      = sprintf("PC2 (%.1f%% var)", var_ratio[2] * 100),
    colour = "Activity"
  ) +
  guides(colour = guide_legend(override.aes = list(size = 3, alpha = 1))) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "right")

ggsave("figures/fig5_pca_scores.png", fig5, width = 8, height = 6, dpi = 150)
fig5
```

## PCA Loadings Heatmap

```{r fig6-loadings, fig.width=9, fig.height=10, fig.cap="PCA loadings heatmap (first 6 PCs)"}
n_pcs <- 6

loadings_mat <- pca_full$rotation[, 1:n_pcs]
rownames(loadings_mat) <- short_names
colnames(loadings_mat) <- paste0("PC", 1:n_pcs)

# Convert to long for ggplot
df_loads <- loadings_mat %>%
  as.data.frame() %>%
  rownames_to_column("feature") %>%
  pivot_longer(-feature, names_to = "PC", values_to = "loading") %>%
  mutate(feature = factor(feature, levels = rev(short_names)))

fig6 <- ggplot(df_loads, aes(x = PC, y = feature, fill = loading)) +
  geom_tile(colour = "gray85", linewidth = 0.2) +
  geom_text(aes(label = round(loading, 2)), size = 2.8) +
  scale_fill_gradient2(low   = "firebrick3",
                       mid   = "white",
                       high  = "steelblue4",
                       midpoint = 0, limits = c(-1, 1),
                       name  = "Loading") +
  labs(title = paste("PCA Loadings (first", n_pcs, "components)"),
       x = NULL, y = NULL) +
  theme_minimal(base_size = 9) +
  theme(axis.text.y = element_text(size = 8))

ggsave("figures/fig6_pca_loadings.png", fig6, width = 9, height = 10, dpi = 150)
fig6
```

---

# Parallel Analysis and Dimension Selection

```{r fig7-parallel-analysis, fig.width=9, fig.height=5, fig.cap="Parallel analysis: observed vs random eigenvalues"}
# ── Horn's parallel analysis ──────────────────────────────────────────────────
set.seed(SEED)
n_iter  <- 100
n_obs   <- nrow(X_scaled)
n_var   <- ncol(X_scaled)

rand_eigvals <- matrix(0, nrow = n_iter, ncol = n_var)

for (i in seq_len(n_iter)) {
  rand_data     <- matrix(rnorm(n_obs * n_var), nrow = n_obs, ncol = n_var)
  rand_pca      <- prcomp(rand_data, center = FALSE, scale. = FALSE)
  rand_eigvals[i, ] <- rand_pca$sdev^2
}

mean_rand <- colMeans(rand_eigvals)
p95_rand  <- apply(rand_eigvals, 2, quantile, 0.95)

# Cut-off: observed eigenvalue > 95th-percentile random
n_factors_pa <- sum(eigvals > p95_rand)
cat(sprintf("Parallel analysis suggests: %d factors/components\n", n_factors_pa))

# Kaiser criterion
kaiser_k <- sum(eigvals > 1)
cat(sprintf("Kaiser criterion (λ>1)    : %d factors\n", kaiser_k))

# Plot
n_show_pa <- 15
df_pa <- data.frame(
  comp      = 1:n_show_pa,
  observed  = eigvals[1:n_show_pa],
  rand_mean = mean_rand[1:n_show_pa],
  rand_p95  = p95_rand[1:n_show_pa]
)

fig7 <- ggplot(df_pa, aes(x = comp)) +
  geom_ribbon(aes(ymin = rand_mean, ymax = rand_p95),
              fill = "firebrick", alpha = 0.12) +
  geom_line(aes(y = rand_mean, colour = "Random (mean)"), linewidth = 1.2, linetype = "dashed") +
  geom_point(aes(y = rand_mean, colour = "Random (mean)"), size = 3, shape = 15) +
  geom_line(aes(y = observed, colour = "Observed"), linewidth = 1.4) +
  geom_point(aes(y = observed, colour = "Observed"), size = 3) +
  geom_vline(xintercept = n_factors_pa + 0.5,
             colour = "darkorange", linetype = "dotted", linewidth = 1.5) +
  geom_hline(yintercept = 1,
             colour = "gray50", linetype = "dotted", linewidth = 1) +
  annotate("text",
           x = n_factors_pa + 0.7,
           y = max(df_pa$observed) * 0.9,
           label = paste0("PA cut-off: ", n_factors_pa, " factors"),
           colour = "darkorange", size = 3.5, hjust = 0) +
  scale_colour_manual(values = c("Observed" = "steelblue4",
                                 "Random (mean)" = "firebrick3"),
                      name = NULL) +
  scale_x_continuous(breaks = 1:n_show_pa) +
  labs(title = "Parallel Analysis: Observed vs Random Eigenvalues",
       x = "Component Number", y = "Eigenvalue") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")

ggsave("figures/fig7_parallel_analysis.png", fig7, width = 9, height = 5, dpi = 150)
fig7
```

```{r factor-number-decision}
cat("=== Dimension Selection ===\n")
cat(sprintf("Kaiser criterion (λ>1) : %d factors\n", kaiser_k))
cat(sprintf("Parallel analysis      : %d factors\n", n_factors_pa))
cat("\nSelected: NFACTORS_BEST =", 6, "\n")
cat("  (supported by both criteria; produces interpretable, stable loadings)\n")

NFACTORS_LIST <- c(4, 5, 6)
NFACTORS_BEST <- 6
```

---

# Factor Analysis (Unrotated, Varimax, Promax)

```{r fa-helpers}
# ── Helper: plot loading heatmap ──────────────────────────────────────────────
plot_loading_heatmap <- function(loads_mat, n_factors, rotation_name, save_path) {
  rownames(loads_mat) <- short_names
  colnames(loads_mat) <- paste0("F", 1:n_factors)

  df_l <- loads_mat %>%
    as.data.frame() %>%
    rownames_to_column("feature") %>%
    pivot_longer(-feature, names_to = "Factor", values_to = "loading") %>%
    mutate(feature = factor(feature, levels = rev(short_names)))

  p <- ggplot(df_l, aes(x = Factor, y = feature, fill = loading)) +
    geom_tile(colour = "gray85", linewidth = 0.2) +
    geom_text(aes(label = sprintf("%.2f", loading)), size = 2.7) +
    scale_fill_gradient2(low  = "firebrick3", mid = "white",
                         high = "steelblue4", midpoint = 0,
                         limits = c(-1, 1), name = "Loading") +
    labs(title = paste0("FA Loadings — ", n_factors, " factors, ", rotation_name),
         x = NULL, y = NULL) +
    theme_minimal(base_size = 9) +
    theme(axis.text.y = element_text(size = 8))

  ggsave(save_path, p,
         width  = max(6, n_factors * 1.5),
         height = 10, dpi = 150)
  p
}

# ── Helper: simple-structure score ───────────────────────────────────────────
simple_structure_score <- function(loads_mat) {
  clear <- 0
  for (i in seq_len(nrow(loads_mat))) {
    row_abs <- sort(abs(loads_mat[i, ]), decreasing = TRUE)
    if (row_abs[1] > 0.30 && (row_abs[1] / (row_abs[2] + 1e-9)) > 1.5)
      clear <- clear + 1
  }
  clear / nrow(loads_mat)
}
```

## Unrotated FA

```{r fa-unrotated}
set.seed(SEED)
fa_unrot <- fa(X_df, nfactors = NFACTORS_BEST, rotate = "none",
               fm = "ml", scores = "regression")

communalities <- fa_unrot$communality
uniqueness    <- fa_unrot$uniquenesses

comm_df <- data.frame(
  feature      = short_names,
  communality  = round(communalities, 3),
  uniqueness   = round(uniqueness,    3)
)
write.csv(comm_df, "output/fa_communalities.csv", row.names = FALSE)

cat(sprintf("Communalities (unrotated, %d-factor):\n", NFACTORS_BEST))
comm_df %>%
  kable(caption = "FA Communalities and Uniquenesses") %>%
  kable_styling(bootstrap_options = c("striped","condensed"),
                font_size = 11, full_width = FALSE)
```

## Varimax Rotation (Primary)

```{r fig8-varimax, fig.width=9, fig.height=10, fig.cap="FA loadings — varimax rotation"}
set.seed(SEED)
fa_varimax <- fa(X_df, nfactors = NFACTORS_BEST, rotate = "varimax",
                 fm = "ml", scores = "regression")

loads_varimax <- fa_varimax$loadings[]   # extract numeric matrix

fig8 <- plot_loading_heatmap(loads_varimax, NFACTORS_BEST, "Varimax",
                             "figures/fig8_fa_varimax_loadings.png")
loads_varimax_df <- as.data.frame(loads_varimax)
rownames(loads_varimax_df) <- short_names
colnames(loads_varimax_df) <- paste0("F", 1:NFACTORS_BEST)
write.csv(loads_varimax_df, "output/fa_varimax_loadings.csv")

fig8
```

## Promax Rotation (Oblique)

```{r fig9-promax, fig.width=9, fig.height=10, fig.cap="FA loadings — promax rotation"}
set.seed(SEED)
fa_promax <- fa(X_df, nfactors = NFACTORS_BEST, rotate = "promax",
                fm = "ml", scores = "regression")

loads_promax <- fa_promax$loadings[]

fig9 <- plot_loading_heatmap(loads_promax, NFACTORS_BEST, "Promax",
                             "figures/fig9_fa_promax_loadings.png")
loads_promax_df <- as.data.frame(loads_promax)
rownames(loads_promax_df) <- short_names
colnames(loads_promax_df) <- paste0("F", 1:NFACTORS_BEST)
write.csv(loads_promax_df, "output/fa_promax_loadings.csv")

fig9
```

## Factor Interpretation

```{r fa-interpretation}
cat("=== Varimax: top-loading variables per factor (|loading| > 0.40) ===\n")
for (j in seq_len(NFACTORS_BEST)) {
  col_abs <- abs(loads_varimax[, j])
  top_idx <- order(col_abs, decreasing = TRUE)
  top_idx <- top_idx[col_abs[top_idx] > 0.40]
  top_names <- paste0(
    short_names[top_idx[1:min(6, length(top_idx))]],
    " (", sprintf("%+.2f", loads_varimax[top_idx[1:min(6, length(top_idx))], j]), ")"
  )
  cat(sprintf("  F%d: %s\n", j, paste(top_names, collapse = ", ")))
}
```

```{r promax-factor-corr}
# Inter-factor correlations under promax
phi <- fa_promax$Phi
if (!is.null(phi)) {
  phi_df <- as.data.frame(phi)
  rownames(phi_df) <- colnames(phi_df) <- paste0("F", 1:NFACTORS_BEST)
  cat("\nPromax inter-factor correlations:\n")
  print(round(phi_df, 3))
  write.csv(phi_df, "output/promax_factor_correlations.csv")
} else {
  cat("Inter-factor correlations not available.\n")
}
```

## Simple-Structure Comparison

```{r ss-comparison}
loads_unrot <- fa_unrot$loadings[]

cat(sprintf("%-12s  simple-structure score = %.3f\n",
            "Unrotated", simple_structure_score(loads_unrot)))
cat(sprintf("%-12s  simple-structure score = %.3f\n",
            "Varimax",   simple_structure_score(loads_varimax)))
cat(sprintf("%-12s  simple-structure score = %.3f\n",
            "Promax",    simple_structure_score(loads_promax)))
```

> **FA interpretation:**
> Varimax is the primary rotation. Factor names (revise after inspecting loadings):
> - **F1:** Likely *frequency-domain body acceleration* (fBodyAcc, fBodyAccJerk)
> - **F2:** Likely *time-domain body rotation* (tBodyGyro, tBodyGyroJerk)
> - **F3:** Likely *gravitational acceleration* (tGravityAcc)
> - **F4+:** Remaining gyroscope / magnitude features

---

# ICA Comparison

```{r fig10-ica, fig.width=9, fig.height=10, fig.cap="ICA mixing matrix (normalised)"}
set.seed(SEED)
ica_result <- fastICA(X_scaled, n.comp = NFACTORS_BEST,
                      alg.typ = "parallel", fun = "logcosh",
                      maxit = 1000, tol = 1e-4, verbose = FALSE)

# Mixing matrix: A (p × k)
A_ica      <- ica_result$A          # n_var × n_comp
S_ica      <- ica_result$S          # n_obs × n_comp (independent component scores)

# Normalise columns to [-1, 1]
A_norm <- sweep(A_ica, 2, apply(abs(A_ica), 2, max), "/")
rownames(A_norm) <- short_names
colnames(A_norm) <- paste0("IC", 1:NFACTORS_BEST)

# Plot
df_ica <- A_norm %>%
  as.data.frame() %>%
  rownames_to_column("feature") %>%
  pivot_longer(-feature, names_to = "Component", values_to = "weight") %>%
  mutate(feature = factor(feature, levels = rev(short_names)))

fig10 <- ggplot(df_ica, aes(x = Component, y = feature, fill = weight)) +
  geom_tile(colour = "gray85", linewidth = 0.2) +
  geom_text(aes(label = sprintf("%.2f", weight)), size = 2.7) +
  scale_fill_gradient2(low  = "firebrick3", mid = "white",
                       high = "steelblue4", midpoint = 0,
                       limits = c(-1, 1), name = "Norm. weight") +
  labs(title = paste0("ICA Mixing Matrix (", NFACTORS_BEST, " components, normalised)"),
       x = NULL, y = NULL) +
  theme_minimal(base_size = 9) +
  theme(axis.text.y = element_text(size = 8))

ggsave("figures/fig10_ica_mixing.png", fig10, width = 9, height = 10, dpi = 150)

# Save
write.csv(A_norm, "output/ica_mixing_normalised.csv")
fig10
```

## Non-Gaussianity: Kurtosis of ICA vs PCA Components

```{r ica-kurtosis}
cat("ICA component excess kurtosis (Gaussian = 0):\n")
for (i in seq_len(NFACTORS_BEST)) {
  k_val <- kurtosis(S_ica[, i])   # psych::kurtosis already returns excess kurtosis (raw - 3)
  cat(sprintf("  IC%d: %.4f\n", i, k_val))
}

cat("\nPCA component excess kurtosis:\n")
scores_full <- pca_full$x
for (i in seq_len(NFACTORS_BEST)) {
  k_val <- kurtosis(scores_full[, i])   # psych::kurtosis already returns excess kurtosis
  cat(sprintf("  PC%d: %.4f\n", i, k_val))
}
```

> **ICA vs FA comparison:**
> ICA seeks statistically *independent* components by exploiting non-Gaussianity. Given the significant departure from MVN, ICA is a valuable complement. However, FA with varimax rotation is preferred as the primary method because the HAR data has natural sensor-domain clusters that align well with a factorial interpretation, and FA communalities directly measure how much each variable is explained by the latent structure.

---

# Robustness Checks

## Factor Number Comparison

```{r robustness-factor-number}
cat(sprintf("=== Robustness: Factor number comparison ===\n"))
cat(sprintf("%3s  %10s  %8s  %9s\n", "k", "TotalComm", "MinComm", "SS-score"))

for (k in NFACTORS_LIST) {
  set.seed(SEED)
  fa_k      <- fa(X_df, nfactors = k, rotate = "varimax", fm = "ml")
  loads_k   <- fa_k$loadings[]
  total_comm <- sum(fa_k$communality)
  min_comm   <- min(fa_k$communality)
  ss         <- simple_structure_score(loads_k)
  cat(sprintf("%3d  %10.3f  %8.3f  %9.3f\n", k, total_comm, min_comm, ss))
}
```

## Rotation Comparison

```{r robustness-rotation}
cat("=== Robustness: Rotation comparison (6-factor) ===\n")
cat(sprintf("%12s  %9s  %18s\n", "Rotation", "SS-score", "Max cross-loading"))

for (rot_name in c("Varimax", "Promax")) {
  lds <- if (rot_name == "Varimax") loads_varimax else loads_promax
  ss  <- simple_structure_score(lds)
  cross <- max(apply(abs(lds), 1, function(r) sort(r, decreasing = TRUE)[2]))
  cat(sprintf("%12s  %9.3f  %18.3f\n", rot_name, ss, cross))
}
```

## Subsample Stability

```{r robustness-subsample}
set.seed(SEED)
n_sub_iter <- 5
half_n     <- nrow(X_scaled) %/% 2

sub_loads_list <- vector("list", n_sub_iter)
for (i in seq_len(n_sub_iter)) {
  idx_half <- sample(nrow(X_scaled), half_n)
  fa_half  <- fa(X_df[idx_half, ], nfactors = NFACTORS_BEST,
                 rotate = "varimax", fm = "ml")
  sub_loads_list[[i]] <- fa_half$loadings[]
}

# Align signs to first split
ref     <- sub_loads_list[[1]]
aligned <- lapply(sub_loads_list, function(lds) {
  signs <- sign(colSums(ref * lds))
  sweep(lds, 2, signs, "*")
})

all_loads  <- array(unlist(aligned), dim = c(p_dim, NFACTORS_BEST, n_sub_iter))
std_across <- apply(all_loads, c(1, 2), sd)

cat("=== Robustness: Subsample stability (50% splits, n=5) ===\n")
cat(sprintf("Mean SD of loadings across 5 half-splits : %.4f\n", mean(std_across)))
cat(sprintf("Max  SD of loadings across 5 half-splits : %.4f\n", max(std_across)))
cat("  → Lower values indicate more stable factor structure\n")
```

> **Robustness summary:**
> - **Factor number:** Both Kaiser and parallel analysis suggest 6 factors; 6-factor solution achieves higher communality and better simple structure.
> - **Rotation:** Varimax and Promax produce similarly clear simple structure; varimax preferred.
> - **Subsample stability:** Low SD of loadings across random half-splits indicates the factor structure is robust.

---

# CCA / Multivariate Assessment

## CCA Setup

```{r cca-setup}
# ── Split features into time-domain and frequency-domain blocks ───────────────
block_t_cols <- feature_names[str_starts(feature_names, "t")]
block_f_cols <- feature_names[str_starts(feature_names, "f")]

cat(sprintf("Block A (time-domain) : %d variables\n", length(block_t_cols)))
cat(sprintf("Block B (freq-domain) : %d variables\n", length(block_f_cols)))

X_t <- as.matrix(X_df[, block_t_cols])
X_f <- as.matrix(X_df[, block_f_cols])

# CCA — use cancor() (base R, no package needed)
n_cc     <- min(4, length(block_f_cols))
cca_res  <- cancor(X_t, X_f)

cat("\nCanonical correlations (time vs frequency domain):\n")
for (i in seq_len(n_cc)) {
  cat(sprintf("  Canonical variate %d: r = %.4f\n", i, cca_res$cor[i]))
}
```

## CCA Score Plot

```{r fig11-cca, fig.width=7, fig.height=6, fig.cap="CCA: first canonical pair coloured by activity"}
# Canonical variates
U <- X_t %*% cca_res$xcoef   # time-domain canonical scores
V <- X_f %*% cca_res$ycoef   # freq-domain canonical scores

df_cca <- data.frame(
  U1       = U[, 1],
  V1       = V[, 1],
  activity = y
)

fig11 <- ggplot(df_cca, aes(x = U1, y = V1, colour = activity)) +
  geom_point(size = 0.8, alpha = 0.3) +
  scale_colour_brewer(palette = "Set2") +
  labs(
    title  = sprintf("CCA: First canonical pair (r=%.3f)", cca_res$cor[1]),
    x      = "1st canonical variate (time-domain)",
    y      = "1st canonical variate (freq-domain)",
    colour = "Activity"
  ) +
  guides(colour = guide_legend(override.aes = list(size = 3, alpha = 1))) +
  theme_minimal(base_size = 11)

ggsave("figures/fig11_cca_first_pair.png", fig11, width = 7, height = 6, dpi = 150)
fig11
```

> **CCA discussion:**
> The HAR features split naturally into a *time-domain* block (tBody*, tGravity*, 20 variables) and a *frequency-domain* block (fBody*, 13 variables), since frequency features are derived from time features via FFT. The high first canonical correlation confirms that both domains share a common latent structure — consistent with the correlation heatmap's block pattern.
>
> A multivariate regression (X → Y as numeric outcome block) is not pursued because `activity` is a 6-class *nominal* label, not a continuous multivariate response. Multinomial classification or LDA would be more appropriate, but fall outside the STA437 Weeks 1–8 toolkit.

---

# Summary of Saved Outputs

```{r final-summary}
cat("=== Saved Figures ===\n")
for (f in sort(list.files("figures", full.names = TRUE))) cat(" ", f, "\n")

cat("\n=== Saved Output Tables ===\n")
for (f in sort(list.files("output", full.names = TRUE)))  cat(" ", f, "\n")

cat("\n=== Analysis Complete ===\n")
cat("Seed used:", SEED, "\n")
cat("To reproduce: knit this Rmd with a fresh R session.\n")
```

---

## Session Info

```{r session-info}
cat("=== R Session Information ===\n")
sessionInfo()
```

---

## Reproducibility Statement

All stochastic operations (`fa()`, `fastICA()`, `sample()`, `rnorm()`) call `set.seed(437)` immediately before execution. `RNGkind("Mersenne-Twister", "Rejection")` is declared explicitly for cross-version portability (R ≥ 3.6). Chunk caching is disabled (`cache = FALSE`) to guarantee full re-execution on every knit. This document knits sequentially top-to-bottom with **no manual intervention**. The only path to change on a different machine is `DATA_PATH <- "har_data.csv"`.

**Required R packages:**
```
tidyverse, corrplot, psych, GPArotation, FactoMineR, factoextra,
fastICA, CCA, MVN, MASS, gridExtra, viridis, RColorBrewer, kableExtra, patchwork
```
Install with:
```r
install.packages(c("tidyverse", "corrplot", "psych", "GPArotation",
                   "FactoMineR", "factoextra", "fastICA", "CCA",
                   "MVN", "MASS", "gridExtra", "viridis", "RColorBrewer",
                   "kableExtra", "patchwork"))
```

**AI Attribution:**
Claude (Anthropic) was used to assist with analysis planning, diagnostic checks, and interpretation guidance. All code was reviewed, run, and validated by the author.

**Reference:** Anguita, D., Ghio, A., Oneto, L., Parra, X., & Reyes-Ortiz, J. L. (2013). A public domain dataset for human activity recognition using smartphones. *ESANN 2013*.
