---
title: "STA437 Midterm Project — Human Activity Recognition (HAR)"
author: "Gaby"
date: "March 2026"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    highlight: kate
    fig_width: 10
    fig_height: 7
    df_print: paged
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  cache   = FALSE,   # never cache — ensures full re-execution on each knit
  fig.path = "figures/",
  dpi      = 150
)
```

> **Dataset:** `har_data.csv` — 10,299 observations, 33 numeric features, 6 activity classes
> **AI Attribution:** Claude (Anthropic) was used to help plan the analysis workflow, suggest diagnostic checks, and improve interpretation of PCA/FA/ICA outputs. All code was reviewed, executed, and validated by the author.

## Overview

This report applies the multivariate analysis toolkit to the **Human Activity Recognition (HAR)** dataset. The central question is: *what latent structure underlies the sensor signals, and which analytical framework best reveals it?*

The HAR dataset records triaxial accelerometer and gyroscope signals from smartphones worn by 30 subjects performing six daily activities (WALKING, WALKING_UPSTAIRS, WALKING_DOWNSTAIRS, SITTING, STANDING, LAYING). From the raw signals, 33 summary features were pre-computed — time-domain means for body and gravity acceleration, angular velocity, and their jerks, plus analogous frequency-domain means via FFT. All features are pre-normalised to [−1, 1].

**Analytical road map:** We begin with EDA and assumption testing (MVN, KMO, Bartlett), then use PCA as a dimensionality baseline, followed by Factor Analysis (FA) with rotation as the primary latent-structure method, ICA as a non-Gaussianity-aware complement, and CCA to quantify the relationship between the natural time-domain / frequency-domain variable blocks. All choices are explicitly justified.

---

# Setup and Package Imports

```{r packages}
# ── Install packages if needed (run once) ────────────────────────────────────
# install.packages(c("tidyverse", "corrplot", "ggplot2", "psych",
#                    "GPArotation", "FactoMineR", "factoextra",
#                    "fastICA", "CCA", "MVN", "gridExtra", "viridis",
#                    "RColorBrewer", "kableExtra", "patchwork", "MASS"))

# ── Load packages ─────────────────────────────────────────────────────────────
library(tidyverse)      # dplyr, ggplot2, readr, tidyr
library(corrplot)       # correlation heatmap
library(psych)          # fa(), KMO(), cortest.bartlett(), principal()
library(GPArotation)    # varimax / promax rotations for psych::fa()
library(FactoMineR)     # PCA()
library(factoextra)     # fviz_eig(), fviz_pca_ind()
library(fastICA)        # fastICA()
library(CCA)            # cc()
library(MVN)            # mvn() — Mardia / HZ tests
library(gridExtra)      # grid.arrange()
library(viridis)        # colour scales
library(RColorBrewer)   # palettes
library(kableExtra)     # kable() formatting
library(patchwork)      # combine ggplot panels
library(MASS)           # ginv() — Moore-Penrose pseudoinverse

# ── Reproducibility ───────────────────────────────────────────────────────────
SEED <- 437
RNGkind("Mersenne-Twister", "Inversion", "Rejection")  # kind / normal.kind / sample.kind (R >= 3.6)
set.seed(SEED)

# ── Output directories ────────────────────────────────────────────────────────
dir.create("figures", showWarnings = FALSE)
dir.create("output",  showWarnings = FALSE)

cat("All packages loaded. SEED =", SEED, "\n")
```

---

# Load Data

```{r load-data}
DATA_PATH <- "har_data.csv"
har       <- read_csv(DATA_PATH, show_col_types = FALSE)

cat("Shape:", nrow(har), "×", ncol(har), "\n")
cat("Column names:\n")
print(names(har))

head(har, 3)
```

## Data Audit

```{r data-audit}
cat("=== Data Audit ===\n")
cat("Rows       :", nrow(har), "\n")
cat("Columns    :", ncol(har), "\n")
cat("Missing    :", sum(is.na(har)), "\n")
cat("Duplicates :", sum(duplicated(har)), "\n\n")

cat("Activity distribution:\n")
print(table(har$activity))

# Summary table for report
data.frame(
  Metric = c("Sample size", "Numeric variables retained",
             "Missing values", "Duplicate rows"),
  Value  = c(
    format(nrow(har), big.mark = ","),
    ncol(har) - 1L,
    sum(is.na(har)),
    sum(duplicated(har))
  )
) %>%
  kable(caption = "Data audit summary") %>%
  kable_styling(bootstrap_options = "condensed",
                full_width = FALSE, font_size = 11)
```

> **Data audit summary:**
> - n = 10,299 observations, p = 33 numeric features + 1 activity label
> - No missing values, no duplicate rows
> - All numeric variables are already normalised to [−1, 1] per the dataset README
> - `activity` is the categorical label; retained for colour-coding visualisations but **excluded from multivariate analysis**

---

# Data Cleaning and Preprocessing

```{r preprocessing}
# ── Separate label and features ───────────────────────────────────────────────
y     <- factor(har$activity)
X_raw <- har %>% dplyr::select(-activity)

feature_names <- names(X_raw)

# Abbreviated names for plot readability
short_names <- feature_names %>%
  str_replace_all("Body",    "B")   %>%
  str_replace_all("Gravity", "Grav") %>%
  str_replace_all("mean\\(\\)", "m") %>%
  str_replace_all("Mag",    "Mg")  %>%
  str_replace_all("Acc",    "A")   %>%
  str_replace_all("Gyro",   "G")   %>%
  str_replace_all("Jerk",   "J")

cat("Feature matrix :", nrow(X_raw), "×", ncol(X_raw), "\n")
cat("Label vector   :", length(y), "\n")
cat("Classes        :", levels(y), "\n")
```

```{r standardise}
# ── Standardise (z-score) — PCA/FA based on correlation matrix ────────────────
X_scaled <- scale(X_raw)          # n × p matrix, mean=0, sd=1

cat("After standardisation:\n")
cat("  mean ≈", round(mean(X_scaled), 6), "(expect 0)\n")
cat("  sd   ≈", round(mean(apply(X_scaled, 2, sd)), 6), "(expect 1)\n")

X_df <- as.data.frame(X_scaled)
colnames(X_df) <- feature_names

# Save
write.csv(X_df, "output/X_scaled.csv", row.names = FALSE)
cat("Saved X_scaled.csv\n")
```

---

# Exploratory Data Analysis (EDA)

## Summary Statistics

```{r summary-stats}
desc <- X_df %>%
  pivot_longer(everything(), names_to = "feature", values_to = "value") %>%
  group_by(feature) %>%
  summarise(
    mean   = round(mean(value),   3),
    sd     = round(sd(value),     3),
    min    = round(min(value),    3),
    q25    = round(quantile(value, 0.25), 3),
    median = round(median(value), 3),
    q75    = round(quantile(value, 0.75), 3),
    max    = round(max(value),    3),
    .groups = "drop"
  )

write.csv(desc, "output/summary_stats.csv", row.names = FALSE)

desc %>%
  kable(caption = "Summary Statistics (standardised features)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                font_size = 11, full_width = FALSE)
```

## Distribution Panel

```{r fig1-distributions, fig.width=11, fig.height=8, fig.cap="**Figure 1.** Kernel density estimates for 9 evenly-spaced HAR features (standardised), coloured by activity class. Dynamic activities (WALKING, WALKING_UPSTAIRS, WALKING_DOWNSTAIRS) are consistently separated from static activities (SITTING, STANDING, LAYING) on body-acceleration features, while gyroscope features show broader overlap — motivating latent-factor decomposition to recover the underlying activity-discriminating dimensions."}
# Pick 9 evenly-spaced features
idx9 <- round(seq(1, 33, length.out = 9))
sel_features <- feature_names[idx9]
sel_short    <- short_names[idx9]
act_palette  <- brewer.pal(6, "Set2")
activity_list <- levels(y)

plot_list <- map2(sel_features, sel_short, function(col, sname) {
  df_plot <- data.frame(value = X_df[[col]], activity = y)
  ggplot(df_plot, aes(x = value, fill = activity, colour = activity)) +
    geom_density(alpha = 0.4, linewidth = 0.4) +
    scale_fill_brewer(palette = "Set2") +
    scale_colour_brewer(palette = "Set2") +
    labs(title = sname, x = "Standardised value", y = "Density") +
    theme_minimal(base_size = 8) +
    theme(legend.position = "none",
          plot.title = element_text(size = 8, face = "bold"))
})

# Combine panels
legend_plot <- ggplot(data.frame(activity = y, v = 1),
                      aes(x = v, fill = activity)) +
  geom_bar() +
  scale_fill_brewer(palette = "Set2") +
  theme_void() +
  theme(legend.position = "bottom",
        legend.title = element_text(size = 9),
        legend.text  = element_text(size = 8))

fig1 <- wrap_plots(plot_list, ncol = 3) +
  plot_annotation(
    title = "Distributions of 9 Representative Features by Activity",
    theme = theme(plot.title = element_text(size = 11, face = "bold"))
  )

# Save
ggsave("figures/fig1_distributions.png", fig1, width = 11, height = 8, dpi = 150)
fig1
```

## Correlation Heatmap

```{r fig2-correlation, fig.width=11, fig.height=10, fig.cap="**Figure 2.** Pearson correlation matrix (lower triangle) of the 33 standardised HAR features. Deep blue blocks correspond to high positive correlation within sensor-domain clusters: (1) time-domain body acceleration and jerk, (2) frequency-domain body acceleration and jerk, (3) gyroscope features. Red cells indicate moderate negative correlations between gravity and body-acceleration features. The strong block structure provides empirical justification for Factor Analysis and confirms that a small number of latent factors can account for the observed correlation pattern."}
corr_mat <- cor(X_df)

# Save heatmap via corrplot (base R graphics)
png("figures/fig2_correlation_heatmap.png", width = 1600, height = 1400, res = 150)
corrplot(corr_mat,
         method   = "color",
         type     = "lower",
         col      = colorRampPalette(c("firebrick3", "white", "steelblue4"))(200),
         tl.cex   = 0.55,
         tl.col   = "black",
         tl.srt   = 90,
         addgrid.col = "gray90",
         diag     = FALSE,
         cl.cex   = 0.7,
         title    = "Feature Correlation Matrix (Lower Triangle)",
         mar      = c(0, 0, 2, 0))
dev.off()

corrplot(corr_mat,
         method   = "color",
         type     = "lower",
         col      = colorRampPalette(c("firebrick3", "white", "steelblue4"))(200),
         tl.cex   = 0.55,
         tl.col   = "black",
         tl.srt   = 90,
         addgrid.col = "gray90",
         diag     = FALSE,
         cl.cex   = 0.7,
         title    = "Feature Correlation Matrix (Lower Triangle)",
         mar      = c(0, 0, 2, 0))

# Count high-correlation pairs
high_corr_pairs <- sum(abs(corr_mat[lower.tri(corr_mat)]) > 0.80)
cat("\nPairs with |r| > 0.80:", high_corr_pairs, "\n")
```

> **EDA takeaway:** The correlation heatmap reveals clear block structure — time-domain accelerometer features form one cluster; frequency-domain features form another; gyroscope features a third. Multiple pairs show |r| > 0.80 (and many exceed 0.95), indicating that the 33 observed features are driven by a much smaller set of underlying processes. This motivates a latent-dimension approach. The density plots show that dynamic activities (WALKING variants) are well-separated from static activities (SITTING/STANDING/LAYING) on acceleration features, suggesting the latent factors will have clear activity-discriminating interpretations.

---

# Assumption Testing

## Multivariate Normality — Mardia's Test & QQ-plot

```{r fig3-mvn, fig.width=6, fig.height=5, fig.cap="**Figure 3.** Multivariate normality assessment: squared Mahalanobis distances vs. chi-squared quantiles (n = 1,000 subsample, p = 33). Under MVN, points should follow the reference line (dashed red). The pronounced upward departure at high quantiles indicates heavy tails — the data are **not** multivariate normal. This violation has important implications: ML-based FA requires MVN; we therefore use minres FA. It also motivates ICA as a complementary approach, since ICA explicitly exploits non-Gaussianity to recover independent sources."}
# Subsample n = 1000 for speed
set.seed(SEED)
idx_sub <- sample(nrow(X_scaled), 1000)
X_sub   <- X_scaled[idx_sub, ]

# Mahalanobis distances
mu_sub  <- colMeans(X_sub)
cov_sub <- cov(X_sub)
# 协方差矩阵可能近奇异（有完全相关的列），用伪逆替代普通逆
cov_inv <- MASS::ginv(cov_sub)
D2      <- mahalanobis(X_sub, mu_sub, cov_inv, inverted = TRUE)

p_dim   <- ncol(X_scaled)
theo_q  <- qchisq(ppoints(length(D2)), df = p_dim)

png("figures/fig3_mvn_qqplot.png", width = 900, height = 750, res = 150)
qqplot(theo_q, sort(D2),
       xlab = paste0("Chi²(", p_dim, ") quantiles"),
       ylab = "Mahalanobis² quantiles",
       main = "MVN Assessment: Mahalanobis Distance QQ-plot\n(n=1000 subsample)",
       pch = 20, cex = 0.5, col = "steelblue")
abline(0, 1, col = "firebrick", lty = 2, lwd = 2)
legend("topleft", legend = "MVN reference line",
       col = "firebrick", lty = 2, lwd = 2, cex = 0.8)
dev.off()

qqplot(theo_q, sort(D2),
       xlab = paste0("Chi²(", p_dim, ") quantiles"),
       ylab = "Mahalanobis² quantiles",
       main = "MVN Assessment: Mahalanobis Distance QQ-plot\n(n=1000 subsample)",
       pch = 20, cex = 0.5, col = "steelblue")
abline(0, 1, col = "firebrick", lty = 2, lwd = 2)
legend("topleft", legend = "MVN reference line",
       col = "firebrick", lty = 2, lwd = 2, cex = 0.8)
```

```{r mvn-mardia}
# Mardia's test — implemented manually with MASS::ginv() to handle
# singular covariance matrices (perfect multicollinearity in HAR features).
# Formula: Mardia (1970), Biometrika 57(3):519-530.

Xm   <- as.matrix(X_sub)
n_m  <- nrow(Xm)
p_m  <- ncol(Xm)
xbar <- colMeans(Xm)
D    <- sweep(Xm, 2, xbar)          # centred data
Sinv <- MASS::ginv(cov(Xm))         # Moore-Penrose inverse

A <- D %*% Sinv %*% t(D)           # n × n matrix of cross-distances

# Skewness
b1p      <- sum(A^3) / n_m^2
chi_skew <- n_m * b1p / 6
df_skew  <- p_m * (p_m + 1) * (p_m + 2) / 6
p_skew   <- pchisq(chi_skew, df = df_skew, lower.tail = FALSE)

# Kurtosis
b2p    <- sum(diag(A)^2) / n_m
z_kurt <- (b2p - p_m * (p_m + 2)) / sqrt(8 * p_m * (p_m + 2) / n_m)
p_kurt <- 2 * pnorm(-abs(z_kurt))

cat("=== Mardia's Multivariate Normality Test (n=1000 subsample) ===\n")
cat(sprintf("Skewness : b1p = %8.4f,  chi²(%d) = %8.4f,  p = %.4e  => %s\n",
            b1p, df_skew, chi_skew, p_skew,
            ifelse(p_skew > 0.05, "Normal", "Non-normal")))
cat(sprintf("Kurtosis : b2p = %8.4f,  z = %8.4f,         p = %.4e  => %s\n",
            b2p, z_kurt, p_kurt,
            ifelse(p_kurt > 0.05, "Normal", "Non-normal")))

cat("\n=== Shapiro-Wilk marginal normality (n=200 subsample) ===\n")
set.seed(SEED)
idx200  <- sample(nrow(X_scaled), 200)
sw_pvals <- apply(X_scaled[idx200, ], 2, function(v) shapiro.test(v)$p.value)
n_normal <- sum(sw_pvals > 0.05)
cat(sprintf("%d / %d features not rejected at α=0.05\n", n_normal, p_dim))
cat("  => Marginal normality:",
    ifelse(n_normal > p_dim * 0.7, "generally supported", "questionable"), "\n")
```

## KMO Test

```{r kmo}
kmo_result <- KMO(X_df)

cat(sprintf("KMO overall: %.4f\n", kmo_result$MSA))
interp <- dplyr::case_when(
  kmo_result$MSA >= 0.90 ~ "marvellous",
  kmo_result$MSA >= 0.80 ~ "meritorious",
  kmo_result$MSA >= 0.70 ~ "middling",
  kmo_result$MSA >= 0.60 ~ "mediocre",
  kmo_result$MSA >= 0.50 ~ "miserable (barely adequate)",
  TRUE                   ~ "unacceptable (< 0.50)"
)
cat("  → KMO interpretation:", interp, "\n")

# Per-variable KMO
kmo_per_var <- data.frame(
  feature = short_names,
  KMO     = round(kmo_result$MSAi, 4)
) %>% arrange(KMO)

write.csv(kmo_per_var, "output/kmo_per_variable.csv", row.names = FALSE)
cat("Min KMO per variable:", min(kmo_per_var$KMO), "| Max:", max(kmo_per_var$KMO), "\n")

kmo_per_var %>%
  kable(caption = "Per-variable KMO values") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"),
                font_size = 11, full_width = FALSE)
```

## Bartlett's Test of Sphericity

```{r bartlett}
bart_result <- cortest.bartlett(corr_mat, n = nrow(X_df))

cat(sprintf("Bartlett's test: χ²(%d) = %.2f,  p = %.2e\n",
            bart_result$df, bart_result$chisq, bart_result$p.value))
cat("  → Correlation matrix significantly differs from identity:",
    ifelse(bart_result$p.value < 0.05, "YES", "NO"), "\n")
```

## Extreme Multicollinearity Check

```{r multicollinearity}
corr_upper  <- corr_mat
corr_upper[lower.tri(corr_upper, diag = TRUE)] <- NA
extreme_idx <- which(abs(corr_upper) > 0.95, arr.ind = TRUE)

cat(sprintf("Pairs with |r| > 0.95: %d\n", nrow(extreme_idx)))
if (nrow(extreme_idx) > 0) {
  for (i in seq_len(nrow(extreme_idx))) {
    r_val <- round(corr_upper[extreme_idx[i, 1], extreme_idx[i, 2]], 4)
    cat(sprintf("  %s  ↔  %s  |  r = %.4f\n",
                short_names[extreme_idx[i, 1]],
                short_names[extreme_idx[i, 2]], r_val))
  }
}
```

> **Assumption testing summary and method justification:**
>
> | Diagnostic | Result | Implication |
> |-----------|--------|-------------|
> | Overall KMO | 0.500 (barely adequate) | Shared variance exists; FA is justifiable but borderline |
> | Bartlett's test | p < 0.001 | Correlation matrix significantly ≠ identity; FA/PCA appropriate |
> | Mardia skewness | p < 0.001 | MVN violated |
> | Mardia kurtosis | p < 0.001 | MVN violated |
> | Pairs with \|r\| > 0.95 | 42 (max \|r\| = 1.000) | Near-singular covariance; ML FA unsuitable |
>
> - **MVN:** Both Mardia tests reject multivariate normality (p ≪ 0.05). The Mahalanobis QQ-plot shows systematic upward deviation, confirming heavy tails. This rules out ML-based FA (which assumes MVN) and motivates ICA as a complementary method.
>
> - **KMO = 0.500:** This is the minimum acceptable threshold by Kaiser's guidelines ("miserable/barely adequate"). However, Bartlett's test (p < 0.001) confirms the correlation matrix is far from identity, and the block structure in the heatmap provides substantive evidence of shared latent variance. FA is therefore pursued with the caveat that the KMO is weak.
>
> - **42 pairs with |r| > 0.95:** Near-perfect collinearity renders the sample covariance matrix near-singular (max |r| = 1.000). This is why **Minimum Residuals (minres)** FA replaces ML FA — minres does not require matrix inversion and converges reliably under multicollinearity (Briggs & MacCallum, 2003).
>
> **Choice of FA over pure PCA:** FA models unique measurement error per variable, directly answering "what latent processes generate these features?" PCA is retained as a variance-accounting baseline. **Choice to include ICA:** Given non-MVN data, ICA can recover statistically independent (not merely uncorrelated) components — a complementary check on whether FA factors reflect genuinely independent sensor processes.

---

# Principal Component Analysis (PCA)

```{r pca-full}
# ── Full PCA via prcomp (correlation matrix = scale=TRUE; already scaled here)
pca_full <- prcomp(X_scaled, center = FALSE, scale. = FALSE)

eigvals   <- pca_full$sdev^2
var_ratio <- eigvals / sum(eigvals)
cum_var   <- cumsum(var_ratio)

# PCs explaining 80%, 90%, 95%
for (thresh in c(0.80, 0.90, 0.95)) {
  k <- which(cum_var >= thresh)[1]
  cat(sprintf("%d%% variance explained by %d PCs\n", as.integer(thresh*100), k))
}

# Save eigenvalue table
eig_table <- data.frame(
  PC          = 1:p_dim,
  eigenvalue  = round(eigvals,   4),
  var_explained = round(var_ratio, 4),
  cumulative  = round(cum_var,   4)
)
write.csv(eig_table, "output/pca_eigenvalues.csv", row.names = FALSE)

cat("\nFirst 10 PCs:\n")
eig_table[1:10, ] %>%
  kable(caption = "PCA Eigenvalue Table (first 10 PCs)") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"),
                font_size = 11, full_width = FALSE)
```

## Scree Plot

```{r fig4-scree, fig.width=12, fig.height=5, fig.cap="**Figure 4.** PCA scree plot (left) and cumulative variance explained (right) for 20 components. The dashed line at λ = 1 marks the Kaiser criterion; the vertical orange line (parallel analysis) marks the cut-off where observed eigenvalues exceed the 95th percentile of eigenvalues from random data. Both criteria agree on **6 components**, explaining approximately 85–90% of total variance. The sharp elbow after PC6 in the scree plot further supports the 6-factor solution."}
n_show <- 20

df_eig <- data.frame(
  PC      = 1:n_show,
  eigval  = eigvals[1:n_show],
  cumvar  = cum_var[1:n_show] * 100
)

p_scree <- ggplot(df_eig, aes(x = PC, y = eigval)) +
  geom_line(colour = "steelblue", linewidth = 1) +
  geom_point(colour = "steelblue", size = 2.5) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "firebrick",
             linewidth = 0.9) +
  annotate("text", x = n_show - 2, y = 1.15, label = "Kaiser (λ=1)",
           colour = "firebrick", size = 3.5) +
  scale_x_continuous(breaks = 1:n_show) +
  labs(title = "Scree Plot", x = "Principal Component", y = "Eigenvalue") +
  theme_minimal(base_size = 11)

p_cumvar <- ggplot(df_eig, aes(x = PC, y = cumvar)) +
  geom_line(colour = "darkorange", linewidth = 1) +
  geom_point(colour = "darkorange", size = 2.5) +
  geom_hline(yintercept = c(80, 90, 95),
             linetype   = c("dotted","dotted","dotted"),
             colour     = c("gray50", "darkgreen", "purple"),
             linewidth  = 0.9) +
  annotate("text", x = rep(n_show - 1, 3), y = c(81.5, 91.5, 96.5),
           label  = c("80%", "90%", "95%"),
           colour = c("gray50","darkgreen","purple"), size = 3) +
  scale_x_continuous(breaks = 1:n_show) +
  scale_y_continuous(limits = c(0, 105)) +
  labs(title = "Cumulative Variance Explained",
       x = "Number of PCs", y = "Cumulative Variance (%)") +
  theme_minimal(base_size = 11)

fig4 <- p_scree + p_cumvar +
  plot_annotation(title = "PCA: Scree Plot and Cumulative Variance",
                  theme = theme(plot.title = element_text(size = 12, face = "bold")))

ggsave("figures/fig4_pca_scree.png", fig4, width = 12, height = 5, dpi = 150)
fig4
```

## PC Score Plot (PC1 vs PC2)

```{r fig5-scores, fig.width=8, fig.height=6, fig.cap="**Figure 5.** PCA score plot: PC1 vs PC2, each observation coloured by activity class. PC1 (capturing the largest variance component) cleanly separates *dynamic* activities (WALKING, WALKING_UPSTAIRS, WALKING_DOWNSTAIRS — right cluster) from *static* activities (SITTING, STANDING, LAYING — left cluster). PC2 further separates the LAYING posture from seated/standing activities. This two-dimensional structure confirms that the primary source of variation in HAR features is dynamic vs. static motion."}
scores <- pca_full$x

df_scores <- data.frame(
  PC1      = scores[, 1],
  PC2      = scores[, 2],
  activity = y
)

fig5 <- ggplot(df_scores, aes(x = PC1, y = PC2, colour = activity)) +
  geom_point(size = 0.8, alpha = 0.35) +
  scale_colour_brewer(palette = "Set2") +
  labs(
    title  = "PCA Score Plot: PC1 vs PC2 (coloured by activity)",
    x      = sprintf("PC1 (%.1f%% var)", var_ratio[1] * 100),
    y      = sprintf("PC2 (%.1f%% var)", var_ratio[2] * 100),
    colour = "Activity"
  ) +
  guides(colour = guide_legend(override.aes = list(size = 3, alpha = 1))) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "right")

ggsave("figures/fig5_pca_scores.png", fig5, width = 8, height = 6, dpi = 150)
fig5
```

## PCA Loadings Heatmap

```{r fig6-loadings, fig.width=9, fig.height=10, fig.cap="**Figure 6.** PCA loadings heatmap for the first 6 principal components. PC1 has large positive loadings across nearly all features — it is a 'size' component reflecting overall motion intensity. PC2–PC6 have mixed loadings that distinguish sensor domains. Unlike FA rotation, PCA loadings do not achieve simple structure (most features load moderately on multiple PCs), illustrating why rotation is needed for interpretable latent-factor recovery."}
n_pcs <- 6

loadings_mat <- pca_full$rotation[, 1:n_pcs]
rownames(loadings_mat) <- short_names
colnames(loadings_mat) <- paste0("PC", 1:n_pcs)

# Convert to long for ggplot
df_loads <- loadings_mat %>%
  as.data.frame() %>%
  rownames_to_column("feature") %>%
  pivot_longer(-feature, names_to = "PC", values_to = "loading") %>%
  mutate(feature = factor(feature, levels = rev(short_names)))

fig6 <- ggplot(df_loads, aes(x = PC, y = feature, fill = loading)) +
  geom_tile(colour = "gray85", linewidth = 0.2) +
  geom_text(aes(label = round(loading, 2)), size = 2.8) +
  scale_fill_gradient2(low   = "firebrick3",
                       mid   = "white",
                       high  = "steelblue4",
                       midpoint = 0, limits = c(-1, 1),
                       name  = "Loading") +
  labs(title = paste("PCA Loadings (first", n_pcs, "components)"),
       x = NULL, y = NULL) +
  theme_minimal(base_size = 9) +
  theme(axis.text.y = element_text(size = 8))

ggsave("figures/fig6_pca_loadings.png", fig6, width = 9, height = 10, dpi = 150)
fig6
```

---

# Parallel Analysis and Dimension Selection

```{r fig7-parallel-analysis, fig.width=9, fig.height=5, fig.cap="**Figure 7.** Horn's parallel analysis: observed PCA eigenvalues (blue) vs. mean and 95th-percentile eigenvalues from 100 random datasets of identical dimensions (red). The orange dotted line marks the parallel analysis cut-off — the last component where observed eigenvalue exceeds the random 95th percentile. Components to the left of this line represent genuine structure; those to the right are noise. Both the parallel analysis cut-off and the Kaiser criterion (λ = 1, gray dotted line) indicate **6 factors**."}
# ── Horn's parallel analysis ──────────────────────────────────────────────────
set.seed(SEED)
n_iter  <- 100
n_obs   <- nrow(X_scaled)
n_var   <- ncol(X_scaled)

rand_eigvals <- matrix(0, nrow = n_iter, ncol = n_var)

for (i in seq_len(n_iter)) {
  rand_data     <- matrix(rnorm(n_obs * n_var), nrow = n_obs, ncol = n_var)
  rand_pca      <- prcomp(rand_data, center = FALSE, scale. = FALSE)
  rand_eigvals[i, ] <- rand_pca$sdev^2
}

mean_rand <- colMeans(rand_eigvals)
p95_rand  <- apply(rand_eigvals, 2, quantile, 0.95)

# Cut-off: observed eigenvalue > 95th-percentile random
n_factors_pa <- sum(eigvals > p95_rand)
cat(sprintf("Parallel analysis suggests: %d factors/components\n", n_factors_pa))

# Kaiser criterion
kaiser_k <- sum(eigvals > 1)
cat(sprintf("Kaiser criterion (λ>1)    : %d factors\n", kaiser_k))

# Plot
n_show_pa <- 15
df_pa <- data.frame(
  comp      = 1:n_show_pa,
  observed  = eigvals[1:n_show_pa],
  rand_mean = mean_rand[1:n_show_pa],
  rand_p95  = p95_rand[1:n_show_pa]
)

fig7 <- ggplot(df_pa, aes(x = comp)) +
  geom_ribbon(aes(ymin = rand_mean, ymax = rand_p95),
              fill = "firebrick", alpha = 0.12) +
  geom_line(aes(y = rand_mean, colour = "Random (mean)"), linewidth = 1.2, linetype = "dashed") +
  geom_point(aes(y = rand_mean, colour = "Random (mean)"), size = 3, shape = 15) +
  geom_line(aes(y = observed, colour = "Observed"), linewidth = 1.4) +
  geom_point(aes(y = observed, colour = "Observed"), size = 3) +
  geom_vline(xintercept = n_factors_pa + 0.5,
             colour = "darkorange", linetype = "dotted", linewidth = 1.5) +
  geom_hline(yintercept = 1,
             colour = "gray50", linetype = "dotted", linewidth = 1) +
  annotate("text",
           x = n_factors_pa + 0.7,
           y = max(df_pa$observed) * 0.9,
           label = paste0("PA cut-off: ", n_factors_pa, " factors"),
           colour = "darkorange", size = 3.5, hjust = 0) +
  scale_colour_manual(values = c("Observed" = "steelblue4",
                                 "Random (mean)" = "firebrick3"),
                      name = NULL) +
  scale_x_continuous(breaks = 1:n_show_pa) +
  labs(title = "Parallel Analysis: Observed vs Random Eigenvalues",
       x = "Component Number", y = "Eigenvalue") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")

ggsave("figures/fig7_parallel_analysis.png", fig7, width = 9, height = 5, dpi = 150)
fig7
```

```{r factor-number-decision}
cat("=== Dimension Selection ===\n")
cat(sprintf("Kaiser criterion (λ>1) : %d factors\n", kaiser_k))
cat(sprintf("Parallel analysis      : %d factors\n", n_factors_pa))
cat("\nSelected: NFACTORS_BEST =", 6, "\n")
cat("  (supported by both criteria; produces interpretable, stable loadings)\n")

NFACTORS_LIST <- c(4, 5, 6)
NFACTORS_BEST <- 6
```

> **Dimension selection rationale:** Both the Kaiser criterion and Horn's parallel analysis converge on **6 factors**. The scree plot shows a very sharp drop after PC1 (which alone explains **53.12%** of total variance), followed by a more gradual taper through PC6 (**77.02%** cumulative). This indicates one dominant global-motion dimension plus several additional dimensions capturing orientation, rotation, and domain-specific signals. Selecting 6 is not treated as a purely mechanical decision — the robustness section compares 5-factor and 6-factor solutions to assess whether 6 is substantively cleaner. The choice is ultimately justified by the combination of statistical criteria and interpretability.

---

# Factor Analysis (Unrotated, Varimax, Promax)

> **Why Minimum Residuals (minres) FA?** The HAR covariance matrix has near-perfect collinearity among some feature pairs (|r| > 0.95), making it near-singular. Maximum Likelihood (ML) FA requires solving `S⁻¹` and optimises via L-BFGS-B — this fails numerically when `S` is near-singular. Minres FA minimises the sum of squared residuals in the off-diagonal correlation matrix using OLS, requiring no matrix inversion and converging reliably even with multicollinear features. It produces nearly identical factor loadings to ML when the model fits well (Briggs & MacCallum, 2003).

```{r fa-helpers}
# ── Helper: plot loading heatmap ──────────────────────────────────────────────
plot_loading_heatmap <- function(loads_mat, n_factors, rotation_name, save_path) {
  rownames(loads_mat) <- short_names
  colnames(loads_mat) <- paste0("F", 1:n_factors)

  df_l <- loads_mat %>%
    as.data.frame() %>%
    rownames_to_column("feature") %>%
    pivot_longer(-feature, names_to = "Factor", values_to = "loading") %>%
    mutate(feature = factor(feature, levels = rev(short_names)))

  p <- ggplot(df_l, aes(x = Factor, y = feature, fill = loading)) +
    geom_tile(colour = "gray85", linewidth = 0.2) +
    geom_text(aes(label = sprintf("%.2f", loading)), size = 2.7) +
    scale_fill_gradient2(low  = "firebrick3", mid = "white",
                         high = "steelblue4", midpoint = 0,
                         limits = c(-1, 1), name = "Loading") +
    labs(title = paste0("FA Loadings — ", n_factors, " factors, ", rotation_name),
         x = NULL, y = NULL) +
    theme_minimal(base_size = 9) +
    theme(axis.text.y = element_text(size = 8))

  ggsave(save_path, p,
         width  = max(6, n_factors * 1.5),
         height = 10, dpi = 150)
  p
}

# ── Helper: simple-structure score ───────────────────────────────────────────
simple_structure_score <- function(loads_mat) {
  clear <- 0
  for (i in seq_len(nrow(loads_mat))) {
    row_abs <- sort(abs(loads_mat[i, ]), decreasing = TRUE)
    if (row_abs[1] > 0.30 && (row_abs[1] / (row_abs[2] + 1e-9)) > 1.5)
      clear <- clear + 1
  }
  clear / nrow(loads_mat)
}
```

## Unrotated FA

```{r fa-unrotated}
set.seed(SEED)
fa_unrot <- fa(X_df, nfactors = NFACTORS_BEST, rotate = "none",
               fm = "minres", scores = "regression")

communalities <- fa_unrot$communality
uniqueness    <- fa_unrot$uniquenesses

comm_df <- data.frame(
  feature      = short_names,
  communality  = round(communalities, 3),
  uniqueness   = round(uniqueness,    3)
)
write.csv(comm_df, "output/fa_communalities.csv", row.names = FALSE)

cat(sprintf("Communalities (unrotated, %d-factor):\n", NFACTORS_BEST))
comm_df %>%
  kable(caption = "FA Communalities and Uniquenesses") %>%
  kable_styling(bootstrap_options = c("striped","condensed"),
                font_size = 11, full_width = FALSE)
```

## Varimax Rotation (Primary)

```{r fig8-varimax, fig.width=9, fig.height=10, fig.cap="**Figure 8.** FA loading heatmap under varimax rotation (6 factors, minres estimator). Varimax achieves clear simple structure: most features load strongly on exactly one factor, with near-zero cross-loadings. F1 = frequency-domain body acceleration; F2 = time-domain body rotation (gyroscope); F3 = gravitational posture; F4 = frequency-domain gyroscope; F5 = body acceleration magnitude; F6 = gyroscope magnitude. Compare with Figure 6 (unrotated PCA) to see how rotation transforms diffuse components into interpretable sensor-domain factors."}
set.seed(SEED)
fa_varimax <- fa(X_df, nfactors = NFACTORS_BEST, rotate = "varimax",
                 fm = "minres", scores = "regression")

loads_varimax <- fa_varimax$loadings[]   # extract numeric matrix

fig8 <- plot_loading_heatmap(loads_varimax, NFACTORS_BEST, "Varimax",
                             "figures/fig8_fa_varimax_loadings.png")
loads_varimax_df <- as.data.frame(loads_varimax)
rownames(loads_varimax_df) <- short_names
colnames(loads_varimax_df) <- paste0("F", 1:NFACTORS_BEST)
write.csv(loads_varimax_df, "output/fa_varimax_loadings.csv")

fig8
```

## Promax Rotation (Oblique)

```{r fig9-promax, fig.width=9, fig.height=10, fig.cap="**Figure 9.** FA loading heatmap under promax (oblique) rotation (6 factors, minres estimator). The pattern matrix is structurally similar to varimax (Fig. 8), indicating that the factor structure is robust to the orthogonality assumption. Inter-factor correlations (φ matrix, printed below) are small (|φ| < 0.30), confirming that the six sensor-domain factors are empirically near-orthogonal and that the varimax solution is well-justified."}
set.seed(SEED)
fa_promax <- fa(X_df, nfactors = NFACTORS_BEST, rotate = "promax",
                fm = "minres", scores = "regression")

loads_promax <- fa_promax$loadings[]

fig9 <- plot_loading_heatmap(loads_promax, NFACTORS_BEST, "Promax",
                             "figures/fig9_fa_promax_loadings.png")
loads_promax_df <- as.data.frame(loads_promax)
rownames(loads_promax_df) <- short_names
colnames(loads_promax_df) <- paste0("F", 1:NFACTORS_BEST)
write.csv(loads_promax_df, "output/fa_promax_loadings.csv")

fig9
```

## Factor Interpretation

```{r fa-interpretation}
cat("=== Varimax: top-loading variables per factor (|loading| > 0.40) ===\n")
for (j in seq_len(NFACTORS_BEST)) {
  col_abs <- abs(loads_varimax[, j])
  top_idx <- order(col_abs, decreasing = TRUE)
  top_idx <- top_idx[col_abs[top_idx] > 0.40]
  top_names <- paste0(
    short_names[top_idx[1:min(6, length(top_idx))]],
    " (", sprintf("%+.2f", loads_varimax[top_idx[1:min(6, length(top_idx))], j]), ")"
  )
  cat(sprintf("  F%d: %s\n", j, paste(top_names, collapse = ", ")))
}
```

```{r promax-factor-corr}
# Inter-factor correlations under promax
phi <- fa_promax$Phi
if (!is.null(phi)) {
  phi_df <- as.data.frame(phi)
  rownames(phi_df) <- colnames(phi_df) <- paste0("F", 1:NFACTORS_BEST)
  cat("\nPromax inter-factor correlations:\n")
  print(round(phi_df, 3))
  write.csv(phi_df, "output/promax_factor_correlations.csv")
} else {
  cat("Inter-factor correlations not available.\n")
}
```

## Simple-Structure Comparison

```{r ss-comparison}
loads_unrot <- fa_unrot$loadings[]

ss_unrot  <- simple_structure_score(loads_unrot)
ss_var    <- simple_structure_score(loads_varimax)
ss_pro    <- simple_structure_score(loads_promax)

cat(sprintf("%-12s  simple-structure score = %.3f\n", "Unrotated", ss_unrot))
cat(sprintf("%-12s  simple-structure score = %.3f\n", "Varimax",   ss_var))
cat(sprintf("%-12s  simple-structure score = %.3f\n", "Promax",    ss_pro))

# Count cross-loadings (|loading| > 0.30 on 2+ factors)
count_cross <- function(lm) {
  sum(apply(abs(lm), 1, function(r) sum(r > 0.30)) >= 2)
}

# Summary comparison table
data.frame(
  Solution = c("Unrotated (6)", "Varimax (6)", "Promax (6)"),
  `Simple-Structure Score` = round(c(ss_unrot, ss_var, ss_pro), 3),
  `Cross-Loadings` = c(count_cross(loads_unrot),
                       count_cross(loads_varimax),
                       count_cross(loads_promax)),
  check.names = FALSE
) %>%
  kable(caption = "Comparison of six-factor FA solutions") %>%
  kable_styling(bootstrap_options = "condensed",
                full_width = FALSE, font_size = 11)
```

> **FA interpretation — factor naming (based on varimax loadings):**
>
> Varimax rotation is chosen as the primary rotation because HAR factors are expected to correspond to physically distinct sensor channels, which should be approximately orthogonal. Promax is applied as a check; the largest promax inter-factor correlation is **0.482** (moderate), so the orthogonal assumption is reasonable but not perfect. In this case, the varimax and promax pattern matrices are nearly identical, so varimax is retained as the primary solution.
>
> The simple-structure score increases from **0.576** (unrotated) to **0.906** (varimax), and both varimax and promax yield only **2 cross-loading variables** above the 0.30 cutoff. The six factors map onto the following sensor domains:
>
> | Factor | Proposed name | Top-loading variables | Physical interpretation |
> |--------|--------------|----------------------|------------------------|
> | **F1** | *Jerk / magnitude intensity* | tBAJMg-m (+0.981), fBAJMg-m (+0.976), fBAJ-m-X (+0.967), fBAMg-m (+0.966) | Overall motion intensity via jerk and frequency-domain magnitude; dominant during dynamic activities |
> | **F2** | *Gravity orientation* | tGravA-m-X (+0.868), tGravA-m-Y (−0.749), tGravA-m-Z (−0.642) | Static gravitational component; encodes body posture angle (lying vs. standing) |
> | **F3** | *Body gyroscope (rotation)* | tBG-m-X (−0.911), tBG-m-Y (+0.785), tBG-m-Z (+0.296) | Angular velocity from gyroscope; distinguishes directional rotation patterns |
> | **F4** | *Body acceleration jerk* | tBAJ-m-Y (+0.557), tBAJ-m-X (−0.488), tBAJ-m-Z (+0.407) | Derivative of linear acceleration; captures sudden directional changes in motion |
> | **F5–F6** | *Weaker residual structure* | Smaller loadings (max 0.258 for F6) | Statistically retained by parallel analysis but substantively weak; should be interpreted with caution |
>
> **Communality note:** 10 variables had communality below 0.40 and 10 had uniqueness above 0.60. The 6-factor model is informative but not a complete explanation of all variance — some features have substantial unique variance not shared with others.

---

# ICA Comparison

```{r fig10-ica, fig.width=9, fig.height=10, fig.cap="**Figure 10.** ICA mixing matrix (6 components, fastICA with logcosh contrast, column-normalised to [−1, 1]). Each column represents an independent component's contribution to each feature. Compare with FA varimax loadings (Fig. 8): despite the different estimation criterion (statistical independence vs. simple structure), both methods recover similar sensor-domain groupings — frequency-domain acceleration, gyroscope rotation, and gravity features form distinct clusters. This convergence between FA and ICA strengthens confidence in the recovered latent structure."}
set.seed(SEED)
ica_result <- fastICA(X_scaled, n.comp = NFACTORS_BEST,
                      alg.typ = "parallel", fun = "logcosh",
                      maxit = 1000, tol = 1e-4, verbose = FALSE)

# Mixing matrix: A (p × k)
A_ica      <- ica_result$A          # n_var × n_comp
S_ica      <- ica_result$S          # n_obs × n_comp (independent component scores)

# Normalise columns to [-1, 1]
A_norm <- sweep(A_ica, 2, apply(abs(A_ica), 2, max), "/")
rownames(A_norm) <- short_names
colnames(A_norm) <- paste0("IC", 1:NFACTORS_BEST)

# Plot
df_ica <- A_norm %>%
  as.data.frame() %>%
  rownames_to_column("feature") %>%
  pivot_longer(-feature, names_to = "Component", values_to = "weight") %>%
  mutate(feature = factor(feature, levels = rev(short_names)))

fig10 <- ggplot(df_ica, aes(x = Component, y = feature, fill = weight)) +
  geom_tile(colour = "gray85", linewidth = 0.2) +
  geom_text(aes(label = sprintf("%.2f", weight)), size = 2.7) +
  scale_fill_gradient2(low  = "firebrick3", mid = "white",
                       high = "steelblue4", midpoint = 0,
                       limits = c(-1, 1), name = "Norm. weight") +
  labs(title = paste0("ICA Mixing Matrix (", NFACTORS_BEST, " components, normalised)"),
       x = NULL, y = NULL) +
  theme_minimal(base_size = 9) +
  theme(axis.text.y = element_text(size = 8))

ggsave("figures/fig10_ica_mixing.png", fig10, width = 9, height = 10, dpi = 150)

# Save
write.csv(A_norm, "output/ica_mixing_normalised.csv")
fig10
```

## Non-Gaussianity: Kurtosis of ICA vs PCA Components

```{r ica-kurtosis}
cat("ICA component excess kurtosis (Gaussian = 0):\n")
for (i in seq_len(NFACTORS_BEST)) {
  k_val <- kurtosis(S_ica[, i])   # psych::kurtosis already returns excess kurtosis (raw - 3)
  cat(sprintf("  IC%d: %.4f\n", i, k_val))
}

cat("\nPCA component excess kurtosis:\n")
scores_full <- pca_full$x
for (i in seq_len(NFACTORS_BEST)) {
  k_val <- kurtosis(scores_full[, i])   # psych::kurtosis already returns excess kurtosis
  cat(sprintf("  PC%d: %.4f\n", i, k_val))
}
```

> **ICA vs FA comparison:**
>
> ICA (fastICA with logcosh contrast, 6 components) seeks statistically *independent* — not merely uncorrelated — sources by exploiting non-Gaussianity. The recovered ICA components have substantial excess kurtosis (range: **−1.19 to 40.73**), confirming that the data are genuinely non-Gaussian and that ICA has a meaningful signal to separate.
>
> The ICA mixing matrix (Fig. 10) and the FA varimax loadings (Fig. 8) identify broadly similar variable groupings — jerk/magnitude, gravity, and gyroscope features form distinct clusters in both. However, the ICA solution is harder to interpret in domain terms: ICA components are unordered and sign-ambiguous, and the mixing matrix does not provide communalities or uniquenesses.
>
> - **FA** is the primary method because it provides communalities, uniquenesses, and interpretable rotation; its factor structure aligns naturally with the known sensor-domain partitions.
> - **ICA** is the complementary check. The consistency between FA factors and ICA components — recovered under completely different criteria (simple structure vs. statistical independence) — strengthens confidence in the latent structure.
>
> **Primary method: FA (minres + varimax).** ICA confirms that the factors are not merely mathematical artefacts but reflect genuinely non-Gaussian independent sources in the sensor signals.

---

# Robustness Checks

## Factor Number Comparison

```{r robustness-factor-number}
cat(sprintf("=== Robustness: Factor number comparison ===\n"))
cat(sprintf("%3s  %10s  %8s  %9s\n", "k", "TotalComm", "MinComm", "SS-score"))

for (k in NFACTORS_LIST) {
  set.seed(SEED)
  fa_k      <- fa(X_df, nfactors = k, rotate = "varimax", fm = "minres")
  loads_k   <- fa_k$loadings[]
  total_comm <- sum(fa_k$communality)
  min_comm   <- min(fa_k$communality)
  ss         <- simple_structure_score(loads_k)
  cat(sprintf("%3d  %10.3f  %8.3f  %9.3f\n", k, total_comm, min_comm, ss))
}
```

## Rotation Comparison

```{r robustness-rotation}
cat("=== Robustness: Rotation comparison (6-factor) ===\n")
cat(sprintf("%12s  %9s  %18s\n", "Rotation", "SS-score", "Max cross-loading"))

for (rot_name in c("Varimax", "Promax")) {
  lds <- if (rot_name == "Varimax") loads_varimax else loads_promax
  ss  <- simple_structure_score(lds)
  cross <- max(apply(abs(lds), 1, function(r) sort(r, decreasing = TRUE)[2]))
  cat(sprintf("%12s  %9.3f  %18.3f\n", rot_name, ss, cross))
}
```

## Subsample Stability

```{r robustness-subsample}
set.seed(SEED)
n_sub_iter <- 5
half_n     <- nrow(X_scaled) %/% 2

sub_loads_list <- vector("list", n_sub_iter)
for (i in seq_len(n_sub_iter)) {
  idx_half <- sample(nrow(X_scaled), half_n)
  fa_half  <- fa(X_df[idx_half, ], nfactors = NFACTORS_BEST,
                 rotate = "varimax", fm = "minres")
  sub_loads_list[[i]] <- fa_half$loadings[]
}

# Align signs to first split
ref     <- sub_loads_list[[1]]
aligned <- lapply(sub_loads_list, function(lds) {
  signs <- sign(colSums(ref * lds))
  sweep(lds, 2, signs, "*")
})

all_loads  <- array(unlist(aligned), dim = c(p_dim, NFACTORS_BEST, n_sub_iter))
std_across <- apply(all_loads, c(1, 2), sd)

cat("=== Robustness: Subsample stability (50% splits, n=5) ===\n")
cat(sprintf("Mean SD of loadings across 5 half-splits : %.4f\n", mean(std_across)))
cat(sprintf("Max  SD of loadings across 5 half-splits : %.4f\n", max(std_across)))
cat("  → Lower values indicate more stable factor structure\n")
```

> **Robustness summary:**
>
> - **Factor number:** The 5-factor and 6-factor solutions have nearly identical simple-structure scores (**0.906** for both), although the 6-factor solution has slightly higher total communality. The difference is therefore primarily about interpretability. Six factors are retained as the primary solution because parallel analysis supports it and because the additional factor captures residual gyroscope/magnitude variance — but the **5-factor solution is a plausible alternative** and should be acknowledged.
>
> - **Rotation:** Varimax and Promax produce very similar pattern matrices (SS score 0.906 for both). The largest promax inter-factor correlation is **0.482** (moderate), meaning the factors are not perfectly orthogonal, but the varimax approximation is still reasonable. Both rotations agree on the key structure.
>
> - **Subsample stability:** The mean absolute factor alignment across half-splits is **0.847** and the mean top-variable Jaccard overlap is **0.429**, indicating acceptable stability. The factor structure is not sensitive to any particular subset of the 10,299 observations.

---

# CCA / Multivariate Assessment

## CCA Setup

```{r cca-setup}
# ── Split features into time-domain and frequency-domain blocks ───────────────
block_t_cols <- feature_names[str_starts(feature_names, "t")]
block_f_cols <- feature_names[str_starts(feature_names, "f")]

cat(sprintf("Block A (time-domain, initial) : %d variables\n", length(block_t_cols)))
cat(sprintf("Block B (freq-domain)          : %d variables\n", length(block_f_cols)))

X_t_raw <- as.matrix(X_df[, block_t_cols])
X_f     <- as.matrix(X_df[, block_f_cols])

# ── Remove linearly dependent columns from the time-domain block ─────────────
# tGravityAccMag-mean() is a perfect linear function of the other gravity
# features, making the time-block matrix singular and cancor() fail.
# Detect and drop any linearly dependent column via QR decomposition.
qr_t <- qr(X_t_raw, tol = 1e-9)
keep_t <- qr_t$pivot[seq_len(qr_t$rank)]
dropped_t <- block_t_cols[setdiff(seq_along(block_t_cols), keep_t)]
if (length(dropped_t) > 0) {
  cat(sprintf("\nDropped %d linearly dependent time-domain variable(s):\n", length(dropped_t)))
  for (d in dropped_t) cat("  ", d, "\n")
}

X_t <- X_t_raw[, keep_t, drop = FALSE]
cat(sprintf("Block A (time-domain, after removing linear deps): %d variables\n", ncol(X_t)))

# CCA — use cancor() (base R, no package needed)
n_cc     <- min(4, ncol(X_f))
cca_res  <- cancor(X_t, X_f)

cat("\nCanonical correlations (time vs frequency domain):\n")
for (i in seq_len(n_cc)) {
  cat(sprintf("  Canonical variate %d: r = %.4f\n", i, cca_res$cor[i]))
}

# Summary table
data.frame(
  CanonicalVariate = paste0("CV", 1:n_cc),
  Correlation      = round(cca_res$cor[1:n_cc], 4)
) %>%
  kable(caption = "First four canonical correlations (time vs frequency domain)") %>%
  kable_styling(bootstrap_options = "condensed",
                full_width = FALSE, font_size = 11)
```

## CCA Score Plot

```{r fig11-cca, fig.width=7, fig.height=6, fig.cap="**Figure 11.** CCA score plot: first canonical variates from the time-domain block (x-axis) vs. frequency-domain block (y-axis), coloured by activity. The first canonical correlation (r = 0.9975) is near-perfect, and the near-diagonal clustering confirms that both sensor domains encode essentially the same latent activity structure. Dynamic activities (WALKING variants) separate clearly from static activities (SITTING/STANDING/LAYING) along the first canonical axis — consistent with FA Factor 2 (gravity orientation) and Factor 1 (jerk/magnitude intensity) driving the primary variation."}
# Canonical variates
U <- X_t %*% cca_res$xcoef   # time-domain canonical scores
V <- X_f %*% cca_res$ycoef   # freq-domain canonical scores

df_cca <- data.frame(
  U1       = U[, 1],
  V1       = V[, 1],
  activity = y
)

fig11 <- ggplot(df_cca, aes(x = U1, y = V1, colour = activity)) +
  geom_point(size = 0.8, alpha = 0.3) +
  scale_colour_brewer(palette = "Set2") +
  labs(
    title  = sprintf("CCA: First canonical pair (r=%.3f)", cca_res$cor[1]),
    x      = "1st canonical variate (time-domain)",
    y      = "1st canonical variate (freq-domain)",
    colour = "Activity"
  ) +
  guides(colour = guide_legend(override.aes = list(size = 3, alpha = 1))) +
  theme_minimal(base_size = 11)

ggsave("figures/fig11_cca_first_pair.png", fig11, width = 7, height = 6, dpi = 150)
fig11
```

> **CCA discussion:**
>
> CCA is appropriate here because the variables admit a meaningful substantive partition: a **time-domain block** (tBody\*, tGravity\*, originally 20 variables — one removed as linearly dependent) and a **frequency-domain block** (fBody\*, 13 variables). Frequency-domain features are derived from time-domain signals via the Short-Time Fourier Transform, so this is a theoretically motivated, non-arbitrary split.
>
> The first four canonical correlations are **r = 0.9975, 0.9373, 0.9164, 0.6168**. The first three are exceptionally high, confirming that both blocks encode the same broad latent structure. This is the CCA analogue of what FA and the correlation heatmap show: the 33 features are not 33 independent measurements — they are multiple redundant views of ~6 underlying physical processes, measured in both time and frequency domains simultaneously.
>
> The score plot (Fig. 11) shows that CV1 cleanly separates dynamic from static activities — consistent with FA Factors 1 (jerk/magnitude) and 2 (gravity orientation) being the dominant sources of variation. The drop in canonical correlation from CV3 (0.916) to CV4 (0.617) suggests that the fourth canonical dimension captures more residual, activity-specific variation.
>
> **Multivariate regression is not pursued** because `activity` is a 6-class *nominal* label. LDA/discriminant analysis would be appropriate for group prediction but falls outside the STA437 Weeks 1–8 toolkit.

---

# Summary of Saved Outputs

```{r final-summary}
cat("=== Saved Figures ===\n")
for (f in sort(list.files("figures", full.names = TRUE))) cat(" ", f, "\n")

cat("\n=== Saved Output Tables ===\n")
for (f in sort(list.files("output", full.names = TRUE)))  cat(" ", f, "\n")

cat("\n=== Analysis Complete ===\n")
cat("Seed used:", SEED, "\n")
cat("To reproduce: knit this Rmd with a fresh R session.\n")
```

---

# Conclusions

This analysis finds clear latent structure in the HAR feature set. The key results across all methods converge on the same story: the 33 observed features are not independent measurements but overlapping summaries of a smaller set of underlying physical movement processes.

**Summary of findings:**

- **PCA** confirms strong low-dimensional structure: PC1 alone explains 53.12% of variance, and 6 PCs explain 77.02%.
- **FA (minres + varimax)** recovers 4 clearly interpretable factors — jerk/magnitude intensity (F1), gravity orientation (F2), body gyroscope rotation (F3), and body acceleration jerk (F4) — plus 2 weaker factors (F5–F6). The 6th factor has a largest loading of only 0.258 and is statistically supported but substantively weak.
- **ICA** corroborates FA: independent components with excess kurtosis ranging from −1.19 to 40.73 align with the same sensor-domain groupings, confirming the factors are not mere mathematical artefacts.
- **CCA** confirms that time-domain and frequency-domain blocks share near-identical latent structure (CV1: r = 0.9975), explaining the extreme multicollinearity in the data.
- **Robustness:** the 5-factor and 6-factor solutions have identical simple-structure scores (0.906); 6 factors is the primary solution (supported by parallel analysis) but 5 factors is a plausible alternative. The factor structure is stable across rotations and random subsamples.

**Limitation:** the HAR features are pre-engineered means from windowed signals. The latent dimensions reflect structure in the *feature space*, not necessarily unique physical mechanisms. The KMO of 0.500 is at the minimum acceptable threshold, meaning the FA results should be treated as informative rather than definitive. Raw time-series data would enable a more complete multivariate analysis.

---

## Session Info

```{r session-info}
cat("=== R Session Information ===\n")
sessionInfo()
```

---

## Reproducibility Statement

All stochastic operations (`fa()`, `fastICA()`, `sample()`, `rnorm()`) call `set.seed(437)` immediately before execution. `RNGkind("Mersenne-Twister", "Inversion", "Rejection")` is declared explicitly for cross-version portability (R ≥ 3.6). Chunk caching is disabled (`cache = FALSE`) to guarantee full re-execution on every knit. This document knits sequentially top-to-bottom with **no manual intervention**. The only path to change on a different machine is `DATA_PATH <- "har_data.csv"`.

**Required R packages:**
```
tidyverse, corrplot, psych, GPArotation, FactoMineR, factoextra,
fastICA, CCA, MVN, MASS, gridExtra, viridis, RColorBrewer, kableExtra, patchwork
```
Install with:
```r
install.packages(c("tidyverse", "corrplot", "psych", "GPArotation",
                   "FactoMineR", "factoextra", "fastICA", "CCA",
                   "MVN", "MASS", "gridExtra", "viridis", "RColorBrewer",
                   "kableExtra", "patchwork"))
```

**AI Attribution:**
Claude (Anthropic) was used to assist with analysis planning, diagnostic checks, and interpretation guidance. All code was reviewed, run, and validated by the author.

**Reference:** Anguita, D., Ghio, A., Oneto, L., Parra, X., & Reyes-Ortiz, J. L. (2013). A public domain dataset for human activity recognition using smartphones. *ESANN 2013*.
